---
title: "covid_scraping"
author: "Matilde Jacobsen"
date: "1st November 2020, updated `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal
I want to analyse how lockdowns of different countries in Europe during the pandemic have changed the positivity rate

# Challenge
I have a data-set of positivity rates within Europe, but I need to find some data about when lockdown has happened for the different countries and transform that into a TRUE/FALSE per week


```{r libraries}

#Installing packages
pacman::p_load(tidyverse, rvest, janitor, readxl, naniar)

#Clean environment
rm(list = ls())
```

#Covid data
I downloaded some tesing data on the following site: https://www.ecdc.europa.eu/en/publications-data/covid-19-testing 
```{r}
#Getting dataset
covid <- read_excel("data-raw/FdYeZGfV.xlsx")
```

## Scrape the data

Next, I tried to scrape the content of a Wikipedia site and extract the HTML table (more accurately this site: https://en.wikipedia.org/wiki/National_responses_to_the_COVID-19_pandemic) by following the guidelines from this website: https://dcl-wrangle.stanford.edu/rvest.html
```{r}
url <- "https://en.wikipedia.org/wiki/National_responses_to_the_COVID-19_pandemic"
# scrape the website

# css_selector <- ".wikitable > tbody:nth-child(2)" #hashed out for knitting
# 
# url %>%  #hashed out knitting
#   read_html() %>% #hashed out for knitting
#   html_node(css = css_selector) %>% #hashed out for knitting
#   html_table() #hashed out for knitting

```

I could not use the above way of scraping for a table the rowspan and colspan are < 1 in some cases, so I googled a solution and found a loop that would allow these spans to be overwritten with the content of the "parent item" in the column or row in which it was merged.
Inspiration to loop: https://stackoverflow.com/questions/57279093/rvest-read-table-with-cells-that-span-multiple-rows

I started out by figuring out how I wanted the table to look like in R, by exploring the header lines and how many columns where needed to fit the data from each row:
```{r}
# get the lines of the table
lines <- url %>%
  read_html() %>%
  html_nodes(xpath="//table[starts-with(@class, 'wikitable')]") %>%
  html_nodes(xpath = 'tbody/tr')

##define the empty table
#getting column names from headers
c1 <- lines[[2]]%>%
    html_children()%>%
    html_text()%>%
    gsub("\n","",.)
c2 <- lines[[3]]%>%
    html_children()%>%
    html_text()%>%
    gsub("\n","",.)
#merging column names so that it matches content of rows
cols <- c("Country / Territory", "Place", "First Lockdown Start date", "First Lockdown End date", "First Lockdown Length (days)", "Second Lockdown Start Date", "Second Lockdown End Date", "Second Lockdown Length (days)", "Level")
#Removing header lists
lines <- rlist::list.remove(lines, c(1:3))
lines <- rlist::list.remove(lines, 128)
#counting length of rows
nrow <- length(lines)
#creating empty data.frame
table <- data.frame(matrix(nrow = nrow, ncol = 9))
#assign column names from list above
colnames(table) <- cols
```

Then I spend ages trying to figure out how this loop works, so that I could adapt it to my situation.
And finally it took for ever to get around all of the formatting, so in the end I had to split up the loop in two, and wrangle one of the rows in the middle of the table for it to work properly:
```{r}
#--------------------------------------------#
##########FIRST PART OF LOOP##################
#--------------------------------------------#

# fill the table until Medina
for(i in 1:96){
  linecontent <- lines[[i]]%>%
    html_children()%>%
    html_text()%>%
    gsub("\n","",.)
  
  #Fix column repetitions
  repetition_col <- lines[[i]]%>%
    html_children()%>%
    html_attr("colspan")%>%
    ifelse(is.na(.),1,.) %>% # if no rowspan, then it is a normal row, not a multiple one
    as.numeric

  #Test repetition loop
  for(j in 1:length(repetition_col)){
    span_col <- repetition_col[j]
    if(span_col > 1){
      linecontent[9] <- linecontent[1]
      linecontent <- c(linecontent[1],linecontent[9],linecontent[2:8])
    } else{
      linecontent <- linecontent
    }
  }  
  # attribute the content to free columns
  colselect <- is.na(table[i,])
  table[i,colselect] <- linecontent
  
  #Fix row repetitions
  repetition_row <- lines[[i]]%>%
    html_children()%>%
    html_attr("rowspan")%>%
    ifelse(is.na(.),1,.) %>% # if no rowspan, then it is a normal row, not a multiple one
    as.numeric
  
# repeat the cells of the multiple rows down
  for(k in 1:length(repetition_row)){
    span_row <- repetition_row[k]
    if(span_row > 1 & colselect[1,1] == FALSE & colselect[1,3] == TRUE){
      table[(i+1):(i+span_row-1),k+1] <- rep(linecontent[k],span_row-1)
    } else if(span_row > 1 & colselect[1,3] == FALSE){
      table[(i+1):(i+span_row-1),k+2] <- rep(linecontent[k],span_row-1)
    } else {
      if(span_row > 1){
      table[(i+1):(i+span_row-1),k] <- rep(linecontent[k],span_row-1)
    }
  }
}
  } 

#---------------------------------#
##########FIX MEDINA##############
#--------------------------------#
#Identifying Medina
medina <- lines[[97]]%>%
    html_children()%>%
    html_text()%>%
    gsub("\n","",.)
#Replace content of length of lockdown
table[97,]$`First Lockdown Length (days)` <- table[97,]$`First Lockdown End date`  
#Replace content of End date
table[97,]$`First Lockdown End date` <- ""

# attribute the content to free columns
colselect <- is.na(table[97,])
table[97,colselect] <- medina

#--------------------------------------------#
##########SECOND PART OF LOOP#################
#--------------------------------------------#  
# fill the table from after Medina
for(i in 98:nrow){
  linecontent <- lines[[i]]%>%
    html_children()%>%
    html_text()%>%
    gsub("\n","",.)
  
  #Fix column repetitions
  repetition_col <- lines[[i]]%>%
    html_children()%>%
    html_attr("colspan")%>%
    ifelse(is.na(.),1,.) %>% # if no rowspan, then it is a normal row, not a multiple one
    as.numeric

  #Test repetition loop
  for(j in 1:length(repetition_col)){
    span_col <- repetition_col[j]
    if(span_col > 1){
      linecontent[9] <- linecontent[1]
      linecontent <- c(linecontent[1],linecontent[9],linecontent[2:8])
    } else{
      linecontent <- linecontent
    }
  }  
  # attribute the content to free columns
  colselect <- is.na(table[i,])
  table[i,colselect] <- linecontent
  
  #Fix row repetitions
  repetition_row <- lines[[i]]%>%
    html_children()%>%
    html_attr("rowspan")%>%
    ifelse(is.na(.),1,.) %>% # if no rowspan, then it is a normal row, not a multiple one
    as.numeric
  
# repeat the cells of the multiple rows down
  for(k in 1:length(repetition_row)){
    span_row <- repetition_row[k]
    if(span_row > 1 & colselect[1,1] == FALSE & colselect[1,3] == TRUE){
      table[(i+1):(i+span_row-1),k+1] <- rep(linecontent[k],span_row-1)
    } else if(span_row > 1 & colselect[1,3] == FALSE){
      table[(i+1):(i+span_row-1),k+2] <- rep(linecontent[k],span_row-1)
    } else {
      if(span_row > 1){
      table[(i+1):(i+span_row-1),k] <- rep(linecontent[k],span_row-1)
    }
  }
}
  } 

```

Now we can have a look at the data
```{r}
head(table)
tail(table)
#display which countries are in the table
levels(unique(as.factor(table$`Country / Territory`)))
#compare to the covid data
levels(unique(as.factor(covid$country)))

#Who has the longest lockdowns in the first period?
table$`First Lockdown Length (days)` <- as.numeric(table$`First Lockdown Length (days)`)
table %>% select(`Country / Territory`,`First Lockdown Start date`,`First Lockdown End date`,`First Lockdown Length (days)`) %>% arrange(desc(`First Lockdown Length (days)`)) %>% head(20)

#Who has the higest positivity rate in the covid data?
covid %>% group_by(country) %>% summarise(sum_of_positivity_rate = sum(positivity_rate)) %>% arrange(desc(sum_of_positivity_rate)) %>% head(20)
```

Next step would now be to create a "week" variable in the table data and merge it with the covid data by week and country and then I would be able to visualize the positivity rating by week and put in information about lockdowns for the individual countries. 
But I will stick to this for now, and maybe finish the script at a later point before handing in the portfolio...
